Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2019-12-30 manuscript/main.Rmd
patch metrics, upper quantile threshold to limit undue influence of very large patches

  df <- df %>%
    filter(area > 1) %>%
    summarize(
      perimeter = paste0(
        round(mean(perimeter)),
        " $\\pm$ ",
        round(sd(perimeter))),
      area = paste0(
        round(mean(area)),
        " $\\pm$ ",
        round(sd(area))),
      ratio = paste0(
        round(mean(pa_ratio),3),
        " $\\pm$ ",
        round(sd(pa_ratio),3)),
      fractal = paste0(
        round(mean(fractal_index),3),
        " $\\pm$ ",
        round(sd(fractal_index),3))
    )

@@

To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape fragmentation metrics [@dale1999]. We report the ratio of edge to area and the fractal index to quantify landscape complexity of forest disturbances. Statistics were calculated for all forest disturbance patches larger than 1 ha using the R package landscapemetrics [@hesselbarth2019]. We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.

@@

Our CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in out-of-sample test data. A visual comparison on recent pan-chromatic Geo-Eye 1 data shows good agreement between the landsat based GFC data and downscaled CNN results (Appendix Figure 7).

[NOTE STATS]

Scaling our classifier to the whole historical orthomosaic we detected 162 km^2^ (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two acquisitions. In total, 99 km^2^, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing, approximate 88 km^2^ (Table1, Figure 5C-D / light green).

Recent deforestation follows a distincly different pattern compared to historical patterns. Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (figure 5D, dark blue and green colours respectively). These differences are reflected in the analysis of landscape metrics of forest loss. Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to 37 ha from 120 ha historically, while perimeter lengths remained approximately the same, with 8653m and 9902m for historical and contemporary maps respectively (Table 2). This shift in perimeter area ratio led to a similar change in the fractal index, increasing in value to 1.13 $\pm$ 0.07 from 1.10 $\pm$ 0.05 over time. Values closer to a fractal index of 2 suggest a more complex (fragmented) landscape.

@@

Changes in both land-use and land-cover led to concomitant changes in above ground carbon stocks. Recovery throughout the region was characterized for patches of forest and plantations. Assuming high density stands, based on previous work, this could amount to a carbon gains of 1592 Gg C, offsetting more recent losses of approximately 1408 Gg C. On the other hand, at the low end, if we assume a lower carbon density of 81.8 Mg C ha^-1^ this would result in a total carbon gain of 811 Gg C. Using our approach results indicate that overall deforestation around Yangambi has resulted in a loss of ~2416 Gg C in AGC stocks.

>>>>

  df <- df %>%
    filter(area > 1,
           area < quantile(area, 0.95)) %>%
    summarize(
      perimeter = paste0(
        round(mean(perimeter),2),
        " $\\pm$ ",
        round(sd(perimeter),2)),
      area = paste0(
        round(mean(area), 2),
        " $\\pm$ ",
        round(sd(area), 2)),
      ratio = paste0(
        round(mean(pa_ratio), 2),
        " $\\pm$ ",
        round(sd(pa_ratio), 2)),
      fractal = paste0(
        round(mean(fractal_index), 2),
        " $\\pm$ ",
        round(sd(fractal_index), 2))
    )

@@

To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape fragmentation metrics [@dale1999]. We report the ratio of edge to area and the fractal index to quantify landscape complexity of forest disturbances. Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package landscapemetrics [@hesselbarth2019]. We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.

@@

Our CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in out-of-sample test data. A visual comparison on recent pan-chromatic Geo-Eye 1 data shows good agreement, for a total of 87% of all pixels, between the landsat based GFC data and downscaled CNN results (Appendix Table 3 and Figure 7). Scaling our classifier to the whole historical orthomosaic we detected 162 km^2^ (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two acquisitions. In total, 99 km^2^, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing, approximate 88 km^2^ (Table1, Figure 5C-D / light green).

Recent deforestation follows a distincly different pattern compared to historical patterns. Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (figure 5D, dark blue and green colours respectively). These differences are reflected in the analysis of landscape metrics of forest loss. Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to ~1.86 $\pm$ 0.75 ha from ~5.25 $\pm$ 5.02 ha historically. Perimeter lengths were longer historically, at 1451.58 $\pm$ 943.27 m,  compared to contemporary landscapes ~921.74 $\pm$ 362.59 m (Table 2). This shift in perimeter area ratio led to a similar change in the fractal index, increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time. Values closer to a fractal index of 2 suggest a more complex (fragmented) landscape.

@@

Changes in both land-use and land-cover led to concomitant changes in above ground carbon stocks. Recovery throughout the region was characterized for patches of forest and plantations. Assuming high density stands, based on previous work, this could amount to a carbon gains of 1592 Gg C across our study area, offsetting more recent losses of approximately 1408 Gg C. On the other hand, at the low end, if we assume a lower carbon density of 81.8 Mg C ha^-1^ this would result in a total carbon gain of 811 Gg C. Using our approach results indicate that overall deforestation around Yangambi has resulted in a loss of ~2416 Gg C in AGC stocks.

### 2019-12-30 manuscript/main.tex
patch metrics, upper quantile threshold to limit undue influence of very large patches

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Land Use and Land Cover change (LULCC) of African rainforest contribute
to global carbon emissions. Yet, most historical estimates of LULCC and
their carbon emissions rely on non-spatially explicity data in the
pre-satellite era (\textless{} 1972). Past studies have expanded this
pre-satellite time frame with declassified satellite surveillance data
or aerial surveys, but none cover the Congo Basin. Here, we use
historical aerial survey photos to map the extent and structure of LULCC
around Yangambi, central Congo Basin, in 1958. Our study leveraged
Structure-from-Motion to generate a large orthomosaic covering 828
km\textsuperscript{2}, counting 733 million pixels, at a sub meter
resolution and geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m.
Primary forest in the orthomosaic was classified with a Convoluted
Neural Network Deep Learning approach. We used these data to quantify
LULCC, landscape and canopy complexity and scale above ground biomass
between historical and contemporary condition. We show a shift from
historical highly structured industrial deforestation in 1958 (162
km\textsuperscript{2}), to contemporary smallholder farming clearing (88
km\textsuperscript{2}), increasing landscape fragmentation and forest
edges. Efforts to quantify canopy texture and their link to carbon
storage had limited to no success. Our analysis provides an insight in
key spatial and temporal patterns of deforestation and reforestation at
a multi-decadal scale and provide a historical context to land-cover and
land-use change spatial patterns for past and ongoing field research in
the area.}

@@

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands,
fields and buildings. We used the Unet Convoluted Neural Net (CNN,
\citet{ronneberger2015}) architecture implemented in Keras
\citep{chollet2015} with an efficientnetb3 backbone
\citep{yakubovskiy2019} running on TensorFlow \citep{martinabadi2015} to
train a binary classifier (i.e.~forest or non-forested). This
methodology is increasingly being used to automate pixel-level
classification in (color) digital photography data \citep{buscombe2018}.
Training data were collected from the orthomosaic by randomly selecting
513 pixel square tiles from homogeneous forested or non-forested areas
within the historical orthomosaic. Homogeneous tiles were combined in
synthetic landscapes using a random gaussian field based binary mask
(Figure 3). We generated 5000 synthetic landscapes for training, while
500 landscapes were generated for both the validation and the testing
dataset. Source tiles did not repeat across datasets to limit
overfitting. In order to limit stitch line misclassifications, along the
seams of mosaicked images, we created synthetic landscapes with
different forest tiles to mimick forest texture transitions. We applied
this technique to 10\% of the generated synthetic landscapes. The CNN
model was trained for 100 epochs on a graphics processing unit (GPU)
maximizing the Intersect-over-Union (IoU) using additional data
augmentation. Data augmentation included random cropping to 320 pixel
squares, random orientation, scaling, perspective, contrast and
brightness shifts and image blurring. During final model evaluation we
report the IoU of our out-of-sample test datasets. The optimized model
was used to classify the complete orthomosaic using a moving window
approach with a step size of 110 pixels and a majority vote across
overlapping areas to limit segmentation edge effects. We refer to Figure
4 for a synoptic overview of the full deep learning learning workflow.

@@

To map long term land-use and land-cover change in the Yangambi region
we used the contemporary Global Forest Change version 1.6 data (GFC,
tile 10N-020E) \citep{hansen2013}. Using this data we calculated the
latest state of the forest with respect to the conditions at the start
of 1958, 60 years earlier. In our analysis we only included forested
pixels which recorded no loss throughout the whole 2000 - 2018 period.
As the resolution of the historical forest classification exceeds that
of the GFC map we downsampled our historical forest cover data to 30m
GFC resolution. We masked out all water bodies using the Global Forest
Change survey data mask layer, and limited the analysis to the right
bank of the Congo river. Where the datasets overlap we provide summary
statistics of deforestation, reforestation and afforestation.

@@

To quantify changes in the structure of forest cover and its
disturbances we used spatial pattern analysis landscape fragmentation
metrics \citep{dale1999}. We report the ratio of edge to area and the
fractal index to quantify landscape complexity of forest disturbances.
Statistics were calculated for all forest disturbance patches larger
than 1 ha using the R package landscapemetrics \citep{hesselbarth2019}.
We provide mean and standard deviation on edge, area, their ratio and
fractal dimension for both the historical and contemporary
\citet{hansen2013} forest cover maps.

@@

We used an area of 400x400m (16 ha) around contemporary permanent
sampling plot locations to quantify the general state of the canopy in
both historical and contemporary remote sensing data. Contemporary data
were resampled to the resolution of the historical data using a nearest
neighbour interpolation for comparison. We used the first acquisition of
a pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado,
U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time). For
permanent sampling plot locations in both historical and contemporary
data a rectangular buffer of 200m around the centre of all plots was
used to create small image subsets. All image subsets were processed
using the R based FOTO implementation and package \citep{hufkens2019a}
using global normalization, ensuring intercomparability of texture
metrics across time and space \citep{Barbier2010}. We used a FOTO
(moving) window size of 187 pixels (or \textasciitilde{}150m), as
described in \citet{solorzano2018} and \citet{Barbier2010}, to ensure
that multiple individual canopies could be included in the analysis. A
buffer with a radius of 50m around the center of all permanent sampling
plots was used to determine the average value of the first principle
component (PC), explaining the majority of the variance in canopy
texture. Where both the Geo-Eye data and the orthomosaic image
overlapped we processed the intersecting region to explore a
wall-to-wall comparison between past and current canopy texture metrics.

@@

Our CNN deep learning classifier reached an Intersection-over-Union
(IoU) accuracy of 97\% on the detection of disturbed forest in
out-of-sample test data. Scaling our classifier to the whole historical
orthomosaic we detected 162 km\textsuperscript{2} (or
\textasciitilde{}20\% of the scene) of disturbed forests. A large
fraction of the disturbed area was restored in the period between the
two acquisitions. In total, 99 km\textsuperscript{2}, or little over
half of the affected forest was restored (Figure 5C-D, dark blue).
Recent deforested areas, as registered through satellite remote sensing,
approximate 88 km\textsuperscript{2} (Table1, Figure 5C-D / light
green).

Recent deforestation follows a distincly different pattern compared to
historical patterns. Historical deforestation showed a classical
fishbone pattern for forest clearing with very sharp edges, while
current patterns are patchy and ad-hoc (figure 5D, dark blue and green
colours respectively). These differences are reflected in the analysis
of landscape metrics of forest loss. Between the historical and
contemporary LULCC maps we see an increase in small disturbances, as
indicated by the decreasing area of the mean patch size, down to 37 ha
from 120 ha historically, while perimeter lengths remained approximately
the same, with 8653m and 9902m for historical and contemporary maps
respectively (Table 2). This shift in perimeter area ratio led to a
similar change in the fractal index, increasing in value to 1.13 \(\pm\)
0.07 from 1.10 \(\pm\) 0.05 over time. Values closer to a fractal index
of 2 suggest a more complex (fragmented) landscape.

@@

Changes in both land-use and land-cover led to concomitant changes in
above ground carbon stocks. Recovery throughout the region was
characterized for patches of forest and plantations. Assuming high
density stands, based on previous work, this could amount to a carbon
gains of 1592 Gg C, offsetting more recent losses of approximately 1408
Gg C. On the other hand, at the low end, if we assume a lower carbon
density of 81.8 Mg C ha\textsuperscript{-1} this would result in a total
carbon gain of 811 Gg C. Using our approach results indicate that
overall deforestation around Yangambi has resulted in a loss of
\textasciitilde{}2416 Gg C in AGC stocks.

@@

\caption{\label{tab:unnamed-chunk-7}Landscape metrics for historical and contemporary deforestation patterns. We report patch perimeter and area, their ratio and fractal dimension. Values are reportedpat as mean $\pm$ standard deviation, across all deforestation patches.}
\centering
\begin{tabular}[t]{lllll}
\toprule
  & perimeter (m) & area (ha) & ratio ($m\textsuperscript{-1}$) & fractal dimension\\
\midrule
historical & 8653 $\pm$ 35676 & 120 $\pm$ 659 & 0.031 $\pm$ 0.014 & 1.099 $\pm$ 0.052\\
contemporary & 9902 $\pm$ 56975 & 37 $\pm$ 226 & 0.046 $\pm$ 0.014 & 1.13 $\pm$ 0.071\\
\bottomrule
\end{tabular}
\end{table}

@@

When classifying our orthomosaic into forest and non-forest states we
favoured a deep learning supervised classification using a Convoluted
Neural Network over manual segmentation to guarantee an
``apples-to-apples'' comparison between the historical and the
contemporary GFC forest cover maps. We acknowledge that both
classifications use different features, i.e.~spectral or spatial data,
but attain a similarly high accuracy of 97\% and 99\%
\citep{hansen2013}, for the GFC and our CNN map respectively. Despite a
number of challenges we were able to map and quantify a vast continuous
area of tropical forest, with limited manual annotation requirements.

@@

All analysis code is available as an R \citep{rcoreteam2019} project
(\url{https://github.com/khufkens/orthodrc}). The analysis relied
heavily on the `raster' \citep{hijmans2019}, `RStoolbox'
\citep{leutner2019}, and `landscapemetrics' \citep{hesselbarth2019}
packages, while post-processing and plotting was facilitated by the
`tidyverse' ecosystem \citep{wickham2017}, `ggthemes'
\citep{arnold2019}, `scales' \citep{wickham2018} and `cowplot'
\citep{wilke2019}. Additional plotting elements were formatted or
provided by `sf' \citep{pebesma2018} and `rnaturalearth'
\citep{south2017} packages, respectively. I'm grateful for the
contributions to the scientific community by the developers of these
packages.

>>>>

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Land Use and Land Cover change (LULCC) of African rainforest contribute
to global carbon emissions. Yet, most historical estimates of LULCC and
their carbon emissions rely on non-spatially explicity data in the
pre-satellite era (\textless{} 1972). Past studies have expanded this
pre-satellite time frame with declassified satellite surveillance data
or aerial surveys, but none cover the Congo Basin. Here, we use
historical aerial survey photos to map the extent and structure of LULCC
around Yangambi, central Congo Basin, in 1958. Our study leveraged
Structure-from-Motion to generate a large orthomosaic covering 828
km\textsuperscript{2}, counting 733 million pixels, at a sub meter
resolution and geo-referenced to \textasciitilde{}4.7 \(\pm\) 4.3 m.
Primary forest in the orthomosaic was classified with a Convolutional
Neural Network Deep Learning approach. We used these data to quantify
LULCC, landscape and canopy complexity and scale above ground biomass
between historical and contemporary condition. We show a shift from
historical highly structured industrial deforestation in 1958 (162
km\textsuperscript{2}), to contemporary smallholder farming clearing (88
km\textsuperscript{2}), increasing landscape fragmentation and forest
edges. Efforts to quantify canopy texture and their link to carbon
storage had limited to no success. Our analysis provides an insight in
key spatial and temporal patterns of deforestation and reforestation at
a multi-decadal scale and provide a historical context to land-cover and
land-use change spatial patterns for past and ongoing field research in
the area.}

@@

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands,
fields and buildings. We used the Unet Convolutional Neural Net (CNN,
\citet{ronneberger2015}) architecture implemented in Keras
\citep{chollet2015} with an efficientnetb3 pre-processing backbone
\citep{yakubovskiy2019} running on TensorFlow \citep{martinabadi2015} to
train a binary classifier (i.e.~forest or non-forested). This
methodology is increasingly being used to automate pixel-level
classification in (color) digital photography data \citep{buscombe2018}.
Training data were collected from the orthomosaic by randomly selecting
513 pixel square tiles from homogeneous forested or non-forested areas
within the historical orthomosaic. Homogeneous tiles were combined in
synthetic landscapes using a random gaussian field based binary mask
(Figure 3). We generated 5000 synthetic landscapes for training, while
500 landscapes were generated for both the validation and the testing
dataset. Source tiles did not repeat across datasets to limit
overfitting. In order to limit stitch line misclassifications, along the
seams of mosaicked images, we created synthetic landscapes with
different forest tiles to mimick forest texture transitions. We applied
this technique to 10\% of the generated synthetic landscapes. The CNN
model was trained for 100 epochs with a batch size of 30 using Adam
optimization \citep{kingma2017}, maximizing the Intersect-over-Union
(IoU) using Sørensen--Dice and categorical cross-entropy loss functions.
Data augmentation included random cropping to 320 pixel squares, random
orientation, scaling, perspective, contrast and brightness shifts and
image blurring. During final model evaluation we report the IoU of our
out-of-sample test datasets. The optimized model was used to classify
the complete orthomosaic using a moving window approach with a step size
of 110 pixels and a majority vote across overlapping areas to limit
segmentation edge effects. We refer to Figure 4 for a synoptic overview
of the full deep learning learning workflow. In addition, we used the
first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair
(GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11
8:55h GMT or 9:55h local time) to classify and visually assess the
robustness of the algorithm on contemporary remote sensing imagery with
the same moving window approach.

@@

To map long term land-use and land-cover change in the Yangambi region
we used the contemporary Global Forest Change version 1.6 data (GFC,
tile 10N-020E) \citep{hansen2013}. Using this data we calculated the
latest state of the forest with respect to the conditions at the start
of 1958, 60 years earlier. In our analysis we only included forested
pixels which recorded no loss throughout the whole 2000 - 2018 period.
As the resolution of the historical forest classification exceeds that
of the GFC map we downsampled our historical forest cover data to 30m
GFC resolution. We masked out all water bodies using the Global Forest
Change survey data mask layer, and limited the analysis to the right
bank of the Congo river. Where the datasets overlap we provide summary
statistics of deforestation, reforestation and afforestation. The same
methodology was used to quantify CNN model performance across the
Geo-Eye 1 scene.

@@

To quantify changes in the structure of forest cover and its
disturbances we used spatial pattern analysis landscape fragmentation
metrics \citep{dale1999}. We report the ratio of edge to area and the
fractal index to quantify landscape complexity of forest disturbances.
Statistics were calculated for all forest disturbance patches larger
than 1 ha and smaller than the 95\textsuperscript{th} percentile of the
patch size distribution using the R package landscapemetrics
\citep{hesselbarth2019}. We provide mean and standard deviation on edge,
area, their ratio and fractal dimension for both the historical and
contemporary \citet{hansen2013} forest cover maps.

@@

We used an area of 400x400m (16 ha) around contemporary permanent
sampling plot locations to quantify the general state of the canopy in
both historical and contemporary pan-chromatic Geo-Eye 1 remote sensing
data (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11
8:55h GMT or 9:55h local time). Geo-Eye data were resampled to the
resolution of the historical data using a nearest neighbour
interpolation for comparison. For permanent sampling plot locations in
both historical and contemporary data a rectangular buffer of 200m
around the centre of all plots was used to create small image subsets.
All image subsets were processed using the R based FOTO implementation
and package \citep{hufkens2019a} using global normalization, ensuring
intercomparability of texture metrics across time and space
\citep{Barbier2010}. We used a FOTO (moving) window size of 187 pixels
(or \textasciitilde{}150m), as described in \citet{solorzano2018} and
\citet{Barbier2010}, to ensure that multiple individual canopies could
be included in the analysis. A buffer with a radius of 50m around the
center of all permanent sampling plots was used to determine the average
value of the first principle component (PC), explaining the majority of
the variance in canopy texture. Where both the Geo-Eye data and the
orthomosaic image overlapped we processed the intersecting region to
explore a wall-to-wall comparison between past and current canopy
texture metrics.

@@

Our CNN deep learning classifier reached an Intersection-over-Union
(IoU) accuracy of 97\% on the detection of disturbed forest in
out-of-sample test data. A visual comparison on recent pan-chromatic
Geo-Eye 1 data shows good agreement, for a total of 87\% of all pixels,
between the landsat based GFC data and downscaled CNN results (Appendix
Table 3 and Figure 7). Scaling our classifier to the whole historical
orthomosaic we detected 162 km\textsuperscript{2} (or
\textasciitilde{}20\% of the scene) of disturbed forests. A large
fraction of the disturbed area was restored in the period between the
two acquisitions. In total, 99 km\textsuperscript{2}, or little over
half of the affected forest was restored (Figure 5C-D, dark blue).
Recent deforested areas, as registered through satellite remote sensing,
approximate 88 km\textsuperscript{2} (Table1, Figure 5C-D / light
green).

Recent deforestation follows a distincly different pattern compared to
historical patterns. Historical deforestation showed a classical
fishbone pattern for forest clearing with very sharp edges, while
current patterns are patchy and ad-hoc (figure 5D, dark blue and green
colours respectively). These differences are reflected in the analysis
of landscape metrics of forest loss. Between the historical and
contemporary LULCC maps we see an increase in small disturbances, as
indicated by the decreasing area of the mean patch size, down to
\textasciitilde{}1.86 \(\pm\) 0.75 ha from \textasciitilde{}5.25 \(\pm\)
5.02 ha historically. Perimeter lengths were longer historically, at
1451.58 \(\pm\) 943.27 m, compared to contemporary landscapes
\textasciitilde{}921.74 \(\pm\) 362.59 m (Table 2). This shift in
perimeter area ratio led to a similar change in the fractal index,
increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over
time. Values closer to a fractal index of 2 suggest a more complex
(fragmented) landscape.

@@

Changes in both land-use and land-cover led to concomitant changes in
above ground carbon stocks. Recovery throughout the region was
characterized for patches of forest and plantations. Assuming high
density stands, based on previous work, this could amount to a carbon
gains of 1592 Gg C across our study area, offsetting more recent losses
of approximately 1408 Gg C. On the other hand, at the low end, if we
assume a lower carbon density of 81.8 Mg C ha\textsuperscript{-1} this
would result in a total carbon gain of 811 Gg C. Using our approach
results indicate that overall deforestation around Yangambi has resulted
in a loss of \textasciitilde{}2416 Gg C in AGC stocks.

@@

\caption{\label{tab:unnamed-chunk-7}Landscape metrics for historical and contemporary deforestation patterns. We report patch perimeter and area, their ratio and fractal dimension. Values are reportedpat as mean $\pm$ standard deviation, across all deforestation patches.}
\centering
\begin{tabular}[t]{lllll}
\toprule
  & perimeter (m) & area (ha) & ratio ($m\textsuperscript{-1}$) & fractal dimension\\
\midrule
historical & 1451.58 $\pm$ 943.27 & 5.25 $\pm$ 5.02 & 0.03 $\pm$ 0.01 & 1.09 $\pm$ 0.04\\
contemporary & 921.74 $\pm$ 362.59 & 1.86 $\pm$ 0.75 & 0.05 $\pm$ 0.01 & 1.1 $\pm$ 0.05\\
\bottomrule
\end{tabular}
\end{table}

@@

When classifying our orthomosaic into forest and non-forest states we
favoured a deep learning supervised classification using a Convolutional
Neural Network over manual segmentation to guarantee an
``apples-to-apples'' comparison between the historical and the
contemporary GFC forest cover maps. We acknowledge that both
classifications use different features, i.e.~spectral or spatial data,
but attain a similarly high accuracy of 97\% and 99\%
\citep{hansen2013}, for the GFC and our CNN map respectively. Despite a
number of challenges we were able to map and quantify a vast continuous
area of tropical forest, with limited manual annotation requirements.

@@

All analysis code is available as an R / python \citep{rcoreteam2019}
projects (\url{https://khufkens.github.io/orthodrc} \&
\url{https://khufkens.github.io/orthodrc_cnn/}). The analysis relied
heavily on the `raster' \citep{hijmans2019}, `RStoolbox'
\citep{leutner2019}, and `landscapemetrics' \citep{hesselbarth2019}
packages, while post-processing and plotting was facilitated by the
`tidyverse' ecosystem \citep{wickham2017}, `ggthemes'
\citep{arnold2019}, `scales' \citep{wickham2018} and `cowplot'
\citep{wilke2019}. Additional plotting elements were formatted or
provided by `sf' \citep{pebesma2018} and `rnaturalearth'
\citep{south2017} packages, respectively. I'm grateful for the
contributions to the scientific community by the developers of these
packages.
